The Leaky Semicolon

### 2021-10-15 paper/intro.tex
standardizing backarrows to be below
Updates

Recently, two denotational models have been proposed that compute sequential
dependencies semantically. \citet{DBLP:conf/esop/PaviottiCPWOB20} defined
Modular Relaxed Dependencies (\MRD), which use event structures to calculate
dependencies for \cXI, targeting the Intermediate Memory Model (\IMM)
\cite{DBLP:journals/pacmpl/PodkopaevLV19}. \citet{DBLP:journals/pacmpl/JagadeesanJR20} defined
Pomsets\footnote{A pomset is a labeled partial order.} with Preconditions (\PwP), which use preconditions and logic to
calculate dependencies for a Java-like language targeting {multicopy-atomic} (\mca{}) hardware, such as
\armeight{} \cite{DBLP:journals/pacmpl/PulteFDFSS18}.  However, neither paper
treated sequential composition as a first-class citizen.  \MRD{} encoded
sequential composition using continuation-passing, and \PwP{} used prefixing,
adding one event at a time on the left.  In both cases, adding an event
requires perfect knowledge of the future.

@@

This model extends naturally to the case of shared-memory concurrency, leading to a \emph{sequentially consistent}
semantics \cite{Lamport:1979:MMC:1311099.1311750}, in which \emph{program order} inside a thread implies
a total \emph{causal order} between read and write events, for example
(where $\SEMI\!$ has higher precedence than $\PAR$):
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
  \PAR
  \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx1}{\DR{x}{1}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx1}{rx1}
      \wki{wx0}{wx1}
      \wki{wx1}{wy2}
      \wki{ry2}{rx1}
    \end{tikzinline}}
\end{gather*}
In general, there will be many such executions, reflecting different
interleavings of the threads.

@@

In such models, the causal order between events is important,
and includes control and data dependencies ($\xpo$), to avoid
paradoxical ``out of thin air'' examples such as:
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \IF r \THEN \PW{y}{1} \FI}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=165,in=15]{wx1}{rx1}
      \po{rx1}{wy1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}
This candidate execution forms a cycle in causal order, so is disallowed,
but this depends crucially on the control dependency
from $(\DR{x}{1})$ to $(\DW{y}{1})$, and the data dependency
from $(\DR{y}{1})$ to $(\DW{x}{1})$. If either is missing, then this execution
is acyclic and hence allowed. For example dropping the control dependency
results in:
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \PW{y}{1}}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=165,in=15]{wx1}{rx1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}

While syntactic dependency calculation suffices for hardware models, it is
not preserved by common compiler optimizations. For example, if we calculate
control dependencies syntactically, then there is a dependency from
$(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate
execution:
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \IF r \THEN \PW{y}{1} \ELSE \PW{y}{1} \FI}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=165,in=15]{wx1}{rx1}
      \po{rx1}{wy1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}
A compiler may lift the assignment $\PW{y}{1}$ out of the conditional,
thus removing the dependency.

>>>>

Recently, two denotational models have been proposed that compute sequential
dependencies semantically. \citet{DBLP:conf/esop/PaviottiCPWOB20} defined
Modular Relaxed Dependencies (\MRD), which use event structures to calculate
dependencies for \cXI, targeting the Intermediate Memory Model (\IMM)
\cite{DBLP:journals/pacmpl/PodkopaevLV19}. \citet{DBLP:journals/pacmpl/JagadeesanJR20} defined
Pomsets with Preconditions (\PwP), which use preconditions and logic to
calculate dependencies for a Java-like language targeting {multicopy-atomic} (\mca{}) hardware, such as
\armeight{} \cite{DBLP:journals/pacmpl/PulteFDFSS18}.  However, neither paper
treated sequential composition as a first-class citizen.  \MRD{} encoded
sequential composition using continuation-passing, and \PwP{} used prefixing,
adding one event at a time on the left.  In both cases, adding an event
requires perfect knowledge of the future.

@@

This model extends naturally to the case of shared-memory concurrency, leading to a \emph{sequentially consistent}
semantics \cite{Lamport:1979:MMC:1311099.1311750}, in which \emph{program order} inside a thread implies
a total \emph{causal order} between read and write events, for example
(where $\SEMI\!$ has higher precedence than $\PAR$):
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
  \PAR
  \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx1}{\DR{x}{1}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx1}{rx1}
      \wki{wx0}{wx1}
      \wki{wx1}{wy2}
      \wki{ry2}{rx1}
    \end{tikzinline}}
\end{gather*}
We can represent such an execution as a labelled partial order, or
\emph{pomset} \cite{GISCHER1988199}.  A program may give rise to many
executions, each reflecting a different interleaving of the threads.

@@

In such models, the causal order between events is important,
and includes control and data dependencies ($\xpo$), to avoid
paradoxical ``out of thin air'' examples such as:
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \IF r \THEN \PW{y}{1} \FI}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=-165,in=-15]{wx1}{rx1}
      \po{rx1}{wy1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}
This candidate execution forms a cycle in causal order, so is disallowed,
but this depends crucially on the control dependency
from $(\DR{x}{1})$ to $(\DW{y}{1})$, and the data dependency
from $(\DR{y}{1})$ to $(\DW{x}{1})$. If either is missing, then this execution
is acyclic and hence allowed. For example dropping the control dependency
results in:
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \PW{y}{1}}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=-165,in=-15]{wx1}{rx1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}

While syntactic dependency calculation suffices for hardware models, it is
not preserved by common compiler optimizations. For example, if we calculate
control dependencies syntactically, then there is a dependency from
$(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate
execution:
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \IF r \THEN \PW{y}{1} \ELSE \PW{y}{1} \FI}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=-165,in=-15]{wx1}{rx1}
      \po{rx1}{wy1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}
A compiler may lift the assignment $\PW{y}{1}$ out of the conditional,
thus removing the dependency.
